MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments \textbf{Supplementary Information}

### 2022-07-07 main/MSstatsPTM_Main.tex
updated spikein section

Table \ref{tab:dataDescription} summarizes the experiments. Two computer simulations had known ground truth, and varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. Details of computer simulations are available in Supplementary Section 2 and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}). \todo{OV: there is a way to cross-reference sections dynamically across files. Check Meena's MSstats paper, or just double-check that all the sections are referenced correctly in the end}

@@

\medskip \noindent {\bf Evaluation:} We expect the spike-in peptides to change corresponding to the values in Figure \ref{fig:benchmark-design}(b). These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4. Mix 1 vs Mix 4 was treated as a negative control, with no expected fold change. Additionally, the background {\it E. coli} lysate peptides were also treated as negative controls and are not expected to change in any comparison. Therefor we evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.

\todo{Here is the place to describe: 'we expect XXX to not change'. Therefore we evaluated the ability of statistical methods to avoid false positives (i.e., specificity). We expect XXX to have fold changes XXX between conditions XXX. We evaluated the accuracy of estimation of fold changes... and the sensitivity of detecting differentially abundant...}  
%
\todo{I think some mixtures are not comparable because the total protein amount is not the same. I believe only in Mix 1 vs 2 and Mix 3 vs 4 are comparable. I do not think it is correct to compare Mix 1 vs 3 and 2 vs 4 directly because they differ in the total amount of the proteome. Or was it normalized based on what we know is the same? If yes, we need to describe in the beginning of 'Evaluation'}  \todo{The unadjusted proteins... estimation... The adjusted proteins... estimation... Positive controls... Negative controls...}

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadjusted true fold change & Adjusted true fold change \\ \hline \hline
%mix1-mix2  & 0                           & -1                        \\ \hline
%mix1-mix3  & 0                           & 1                         \\ \hline
%mix1-mix4  & -1                          & 0                         \\ \hline
%mix2-mix3  & -1                          & -2                        \\ \hline
%mix2-mix4  & 0                           & -1                        \\ \hline
%mix3-mix4  & 0                           & 1                         \\ \hline
%\end{tabular}
%\end{table}

\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{images/benchmark_fig.png}
\caption{{\bf Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}. (a) Relative amounts of human lysate, background {\it E. Coli} lysate, and spike-in peptides. \todo{TODO: Clarify what is changing how} (b) The expected fold change of the spike-in peptides, after adjusting for changes in the unmodified peptide (human lysate). The expected fold change is calculated using the values of the spike-in and human lysate in part (a).\todo{Could you make the columns in the table more complete? True fold-fold changes between the backgrounds; unadjusted true fold changes, adjusted true fold changes - essentially any value for any comparison that we may want to evaluate}}

>>>>

Table \ref{tab:dataDescription} summarizes the experiments. Two computer simulations had known ground truth, and varied in experimental realism. The first simulation produced a perfectly clean dataset, with many replicates and no missing values. The second simulation introduced real-world characteristics, such as limited modified features and missing values. Details of computer simulations are available in Supplementary Section \ref{supp-sec:sim} and on GitHub (\url{https://github.com/devonjkohler/MSstatsPTM_simulations}).

@@

\medskip \noindent {\bf Evaluation:} We expect the spike-in peptides to change corresponding to the values in Figure \ref{fig:benchmark-design}(b). These peptides are treated as positive controls in all comparisons except Mix 1 vs Mix 4, which was treated as a negative control. Additionally, the background {\it E. coli} lysate peptides were treated as negative controls and were not expected to change in any comparison. We evaluated the statistical methods ability to avoid false positives, as well as their sensitivity in detecting the differentially abundant spike-in peptides and accurately estimate their expected fold change.

@@

%\begin{table}[]
%\begin{tabular}{|c|c|c|}
%\hline
%Comparison & Unadj. true FC & Adj. true FC \\ \hline \hline
%mix1-mix2  & -1                           & 1                        \\ \hline
%mix1-mix3  & 0                           & -1                         \\ \hline
%mix1-mix4  & -1                          & 0                         \\ \hline
%mix2-mix3  & 1                          & -2                        \\ \hline
%mix2-mix4  & 0                           & -1                        \\ \hline
%mix3-mix4  & -1                           & 1                         \\ \hline
%\end{tabular}
%\end{table}

\begin{figure}[ht]
\centering
\includegraphics[scale=.5]{images/benchmark_fig.png}
\caption{{\bf Dataset 3: SpikeIn benchmark - Ubiquitination - Label-free}. (a) Relative amounts of human lysate, background {\it E. Coli} lysate, and spike-in peptides. Volumes of each mixture totaled 20mg, with 10mg of background {\it E. Coli} lysate used to normalize mix 3 and mix 4 to make all mixtures directly comparable. 2 pmol of spike-in peptides was added to mix 1 and mix 3, while 1 pmol was added to mix 2 and mix 4. (b) The expected fold change of the spike-in peptides in each comparison. ``Unadj. true FC'' is the fold change of the spike-in peptides (red) without adjusting for changes in human lysate (blue). This was calculated by directly comparing the amounts of spike-in in each condition. ``Adj. true FC'' is the fold change of the spike-in peptides when adjusting for changes in human lysate. This was calculated by determining the ratio of spike-in peptides to human lysate in each mixture and then using the ratio to calculate the true fold change across comparisons.}

### 2022-07-07 supplementary/ptm_sm.tex
updated spikein section

The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis. To make the calculation the standard deviation and degrees of freedom are needed. The estimation of these quantities is dependent on the experimental design. In Figure \ref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.

@@

The standard error and degrees of freedom are estimated separately for both the PTM and unmodified protein models. Once this is done we can use these these values to calculate the sample size and perform power analysis. Given $q$, the desired false discovery rate, $\beta$, the average Type II error rate, $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison. The minimal number of replicates is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
\[
\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
\]
where 
\[
\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
\]
and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution. The degrees of freedom are estimated for the appropriate model using restricted maximum likelihood.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Computer simulation}

Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance. Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions. To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).

>>>>

%The proposed approach applied to experiments with complex designs still allows for sample size calculation and power analysis. To make the calculation the standard deviation and degrees of freedom are needed. The estimation of these quantities is dependent on the experimental design. In Figure \ref{fig:statistical_inference} different experimental designs that the proposed approach can be applied to, along with their estimation of standard error and degrees of freedom is shown.

@@

%The standard error and degrees of freedom are estimated separately for both the PTM and unmodified protein models. Once this is done we can use these these values to calculate the sample size and perform power analysis. Given $q$, the desired false discovery rate, $\beta$, the average Type II error rate, $m_0 / (m_0 + m_1)$, the fraction of truly differentially modified PTM sites in the comparison. The minimal number of replicates is determined by the variance of the estimated log-fold change $\mathrm{SE}^{2}(\hat{\Delta})$ as
%\[
%\mathrm{SE}^{2}(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]
%\leq \left( \frac{\Delta}{t_{1-\beta, df} + t_{1-\alpha /2, df}} \right)^{2},
%\]
%Where 
%\[
%\alpha = (1 - \beta) \cdot \frac{q}{1 + (1-q) \cdot m_0 / m_1},
%\]
%and $t_{1-\beta, df}$ and $t_{1-\alpha /2, df}$ are the $100(1-\beta)^{\text{th}}$ and the $100(1-\alpha /2)^{\text{th}}$ percentiles of the $t$ distribution. The degrees of freedom are estimated for the appropriate model using restricted maximum likelihood.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Computer simulation}
\label{sec:comp_sim}
Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance. Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions. To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).
